Calculating Biological Module Enrichment or Depletion and Visualizing Data on Large-scale Molecular Maps with \pkg{ACSNMineR} and \pkg{RNaviCell} packages

### 2016-08-23 jss/Deveau.tex
[skip ci] update eric affiliation
[skip ci] update summary and perspectives

\section[Summary and perspectives]{Summary and perspectives}
In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
the calculation of p-values for enrichment or depletion of genes in biological
pathways. The package includes the six large-scale molecular maps and 55
functional modules of the Atlas of Cancer Signaling Network (ACSN) . Enrichment can be
calculated for those maps and modules with several options to play with, but can
also be calculated for other databases of molecular pathways, that can be
imported from GMT formated files.

@@

Of course, \pkg{ACSNMineR} is not the only R package for enrichment
calculations. For instance, \BIOpkg{GOstats} \citep{falcon2007using} is probably
one of the first packages that was created to calculate enrichment for Gene
Ontology categories.  \pkg{GOstats} can also be used to calculate enrichment for
other biological pathways categories, such as KEGG pathways (by using an
instance of the class \code{KEGGHyperGParams}) or PFAM protein families (using
\code{PFAMHyperGParams}). However, its usage might not be as straightforward as
\pkg{ACSNMineR}, and it does not seem possible to test user-defined biological
pathways. Furthermore, other authors have pointed out that the KEGG database
used by this package has not been updated since 2012. \BIOpkg{clusterProfiler} is a
recent R package released for enrichment analysis of Gene Ontology and KEGG with
either hypergeometric test or Gene Set Enrichment Analysis (GSEA)
\citep{yu2012clusterprofiler}. Via other packages, support for analysis of
Disease Ontology and Reactome Pathways is possible. Interestingly, this package
also offers the possibility to import user-defined gene set, through
tab-delimited pairwise definition files.

@@

@@

\bibliography{ACSNMineR}

@@

\address{
  Andrei Zinovyev\\
  Computational Systems Biology of Cancer, Institut Curie\\
  26, rue d'Ulm 75248 Paris\\
  France\\
}
\email{andrei.Zinovyev@curie.fr}

\address{
  Eric Bonnet\\
  Computational Systems Biology of Cancer, Institut Curie\\
  26, rue d'Ulm 75248 Paris\\
  France\\
}
\email{eric.bonnet@curie.fr}

>>>>

\section[Summary and perspectives]{Summary and perspectives}

In this work, we presented the R package \pkg{ACSNMineR}, a novel package for
the calculation of p-values for enrichment or depletion of genes in biological
pathways. The package includes the six large-scale molecular maps and 55
functional modules of the Atlas of Cancer Signaling Network (ACSN) . Enrichment
can be calculated for those maps and modules with several options to play with,
but can also be calculated for other databases of molecular pathways, that can
be imported from GMT formated files.

@@

Of course, \pkg{ACSNMineR} is not the only R package for enrichment
calculations. For instance, \BIOpkg{GOstats} \citep{falcon2007using} is
probably one of the first packages that was created to calculate enrichment for
Gene Ontology categories.  \pkg{GOstats} can also be used to calculate
enrichment for other biological pathways categories, such as KEGG pathways (by
using an instance of the class \code{KEGGHyperGParams}) or PFAM protein
families (using \code{PFAMHyperGParams}). However, its usage might not be as
straightforward as \pkg{ACSNMineR}, and it does not seem possible to test
user-defined biological pathways. Furthermore, other authors have pointed out
that the KEGG database used by this package has not been updated since 2012.
\BIOpkg{clusterProfiler} is a recent R package released for enrichment analysis
of Gene Ontology and KEGG with either hypergeometric test or Gene Set
Enrichment Analysis (GSEA) \citep{yu2012clusterprofiler}. Via other packages,
support for analysis of Disease Ontology and Reactome Pathways is possible.
Interestingly, this package also offers the possibility to import user-defined
gene set, through tab-delimited pairwise definition files. Other notable
packages for enrichment calculations are \pkg{ReactomePA} for Reactome molecular
pathways \citep{ReactomePA2016}, \pkg{miRNApath} for microRNA pathways
\citep{miRNApath2008} and \pkg{gage} \citep{gage2009}. We believe that the main
advantages of ACSNMineR compared to other packages are a direct access to the
full set of ACSN maps (updated on a regular basis) and an easy way to test
MSigDB gene sets or any user-defined gene set formatted appropriately.

@@

\pkg{RNaviCell} relies on standard HTTP calls to provide informations and
calculations, and we have developped a number of bindings for popular
programming languages such as R, Java and Python \citep{bonnet2015navicell}.
This open architecture is designed to facilitate the development of utilities
by other programmers and to facilitate the integration of ACSN maps in existing
frameworks. The development of such services, sometimes called ``microservices"
\citep{fowler2014} is in expansion. Furthermore, this kind of open architecture
could clear the way for a more unified and general access to reaction networks
database, including for example WikiPathways \citep{kelder2012wikipathways},
Reactome \citep{croft2014reactome} and other databases. The PSICQUIC project is
a successfull example of such an architecture \citep{aranda2011psicquic}. It is
an effort of the HUPO proteomics standard initiative to standardize the access
to molecular interaction databases programmatically, based on the specification
of web services (using REST and SOAP calls) and a common query language (MIQL).

@@

\bibliography{Deveau}

@@

\address{
  Andrei Zinovyev\\
  Computational Systems Biology of Cancer, Institut Curie\\
  26, rue d'Ulm 75248 Paris\\
  France\\
}
\email{andrei.zinovyev@curie.fr}

\address{
  Eric Bonnet\\
  Centre National de Génotypage, Institut de Génomique, CEA\\
  2, rue Gaston Crémieux, 91057 Evry\\
  France\\
}
\email{eric.bonnet@cng.fr}
